Verifying a Postfix Language \\ \large CS595 Final Project

### 2021-12-05 project/paper.tex
after presentation
expand proofs

\subsection{Type Identifiers}
Type identifiers use the same store as used for values, however their identifiers are guaranteed to not conflict as type identifiers start with a capitol letter vs lower-case for value identifiers. Because we use the same store the step semantics for closures and scopes continue to work.
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{Type-Let}
            {\tau \prim
                \\ \sigma = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n] ]
                \\ \sigma' = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n, T_{n+1} \tstep \tau] ]}
            {\sigma; \tau ~ \kweid{T} = \tstep \sigma'; \varepsilon}

@@

        % fixed point combinator
        \and \Rule{Fixed Point}{e \val \\ \tau \prim}{\kwfix{e}{\tau} \val}

        % input types for macros can step
        \and \Rule{Macro Type Step}
            {\tau^* \tstep \tau^{*\prime}}
            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{*\prime}}{e^*}}

@@

        \and \Rule{Identifier}
            {\ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{\kwid{x}}{\tau}}
        \and \Rule{Let}
            {\typed{\ctx}{e}{\tau}
                \\ \ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{e~ \kwid{x} =}{\varepsilon}}

@@

            Otherwise, $e_f \val$. To handwave a bit, by inversion on \Rulename{Macro}, $e_f$ is a macro with a body $e^*_b$, and by \Rulename{CALL-1}, $e^* \step e^*_b ~ \textbf{end_scope}$.
    \end{itemize}
\end{proof}

\subsection{Theorem - Preservation}
If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^{*\prime} \tstep^* \tau^*$
% \subsubsection{Lemma - Convergence}
% Part 1: $\tau^* \tstep \tau^{*\prime}$ is no longer possible when $\tau_i \prim ~ \forall~ \tau_i \in \tau^*$.
% \begin{proof}
% Proof by exhaustion with all rules involving $\prim$ types not stepping. To values
% \end{proof}
% Part 2: $\tau^* \tstep \tau^{*\prime}$ will eventually step to a $\tau^{*\prime}$ such that $\tau_i \prim ~ \forall~ \tau_i \in \tau^{*\prime}$.
% \begin{proof}
% This is provable by the fact that the language does not feature recursive types and the only form of recursion in the language is via the fix operator and $\kwType{}$ operator does no have to follow recursive paths in order to get it's type (Type Rules \Rulename{Macro}, \Rulename{Fix}). The other rules are trivial (ie - step to a $\prim$) and/or irrelevant to the assertion
% \end{proof}

>>>>

\subsection{Type Identifiers}
Type identifiers use the same store as used for values, however their identifiers are guaranteed to not conflict as type identifiers start with a capital letter vs lower-case for value identifiers. Because we use the same store the step semantics for closures and scopes continue to work.
{
    \centering
    \def \MathparLineskip{\lineskip=0.43cm}
    \begin{mathpar}
        \Rule{Type-Let}
            {\tau \prim
                \\ \sigma = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n] ]
                \\ \sigma' = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n, T_{n+1} \tstep \tau] ]}
            {\sigma; \tau ~ \kweid{T} = \tstep \sigma'; \varepsilon}

@@

        % fixed point combinator
        \and \Rule{Fixed Point}
            {\kwmacro{\tau^*~ \tau_1}{e^*} \val \\ \tau_1 \prim}
            {\kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \val}
        \and \Rule{Call FP}
            {v = \kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \\ v \val}
            {v~ \textbf{@} \step v~ \kwmacro{\tau^*~ \tau_1}{e^*}~ \textbf{@}}

        % input types for macros can step
        \and \Rule{Macro Type Step}
            {\tau^* \tstep \tau^{*\prime}}
            {\sigma; \kwmacro{\tau^*}{e^*} \tstep \sigma; \kwmacro{\tau^{*\prime}}{e^*}}

@@

        \and \Rule{Identifier}
            {\ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{\kwid{x}}{\tau}}

        \and \Rule{Let}
            {\typed{\ctx}{e}{\tau}
                \\ \ctx(\kwid{x}) = \tau}
            {\typed{\ctx}{e~ \kwid{x} =}{\varepsilon}}

@@

            Otherwise, $e_f \val$. To handwave a bit, by inversion on \Rulename{Macro}, $e_f$ is a macro with a body $e^*_b$, and by \Rulename{CALL-1}, $e^* \step e^*_b ~ \textbf{end_scope}$.

        \item \Rulename{Identifier}
            Then $e^* = \texttt{x}$.
            By, uh, the other \Rulename{Identifier}, $e^*$ steps to a value in the store.

        \item \Rulename{Let}
            Then $e^* = e ~ \texttt{x} =$.
            By induction, $e \val$ or $e \step e'$.
            If $e \step e'$, then by \Rulename{STEP}, $e^* \step e' ~ \texttt{x} =$.
            If $e \val$, then by the other \Rulename{LET}, $\sigma;e^* \step \sigma';\varepsilon$, where $\sigma'$ is, as described in the rule, $\sigma$ with $x \step e$ added to its ``topmost'' scope.

        \item \Rulename{Fix}
            Then $e^* = \kwfix{e}{\tau_1}$.
            By induction, $e \val$ or $e \step e'$.
            If $e \step e'$, then by \Rulename{STEP}, $e^* \step \kwfix{e'}{\tau_1}$.
            If $e \val$, then I think I'm missing something here, but by \Rulename{Fixed Point}, $e^* \val$.
    \end{itemize}
\end{proof}

\subsection{Theorem - Preservation}
If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^* \tstep^* \tau^{*\prime}$.

\begin{proof}
By induction on the derivation of $e^* \step e^{*\prime}$.
\begin{itemize}
    \item \Rulename{STEP}
        Then $e^* = v^* e_1 ... e_n$ and $e^{*\prime} \step v^* e_1' ... e_n$ where $\etyped{e_1}{\tau}$ and $\etyped{e_1'}{\tau'}$.
        By induction, $\tau \tstep^* \tau'$, so by \Rulename{TSTEP}, $\tau^* \tstep^* [\tau' / \tau]\tau^*$.

    \item \Rulename{TUPLES-1}
        Then $e^* = (e_{elem}^*)$ and $e^* \step (e_{elem}^{*\prime})$ and $\tau^* = (\tau_{elem}^*)$.
        By induction, $\etyped{e_{elem}^{*\prime}}{\tau_{elem}^{*\prime}}$ where $\tau_{elem}^* \tstep^* \tau_{elem}^{*\prime}$. By \Rulename{Tuple Type}, $\etyped{(e_{elem}^{*\prime})}{(\tau_{elem}^{*'})}$, and by \Rulename{Tuple Types - 1}, $(\tau_{elem}^*) \tstep (\tau_{elem}^{*'})$.

    \item \Rulename{LET}
        Then $e^* = v ~ \texttt{\$x} ~ =$. By the other \Rulename{Let}, $\etyped{e^*}{\varepsilon}$, so $\tau^* = \varepsilon$. $e^* \step \varepsilon$, so $\tau^{*\prime} = \varepsilon$, so $\tau^* = \tau^{*\prime}$.

    \item \Rulename{Identifier}
        Then $e^* = \texttt{x}$, where $\texttt{x} \step v$ is in scope. By the other \Rulename{Identifier}, the type of an identifier expression is the same as the type of the value in the store ($\etyped{v}{\tau}$), so $\tau^* = \tau^{*\prime} = \tau$.

    \item \Rulename{CALL-1}
        Then $e^* = $.
        % I think we need to modify (CALL-1) to require the values? Revisit.

    \item \Rulename{CALL-2}
        % do we need a type rule for end_scope?

    \item \Rulename{TUPLES-3}
        Then $e^* = \kwunpack{(v^*)}$, where $\etyped{v^*}{\tau_{elem}^*}$.
        By \Rulename{Unpack}, $\tau^* = \tau_{elem}^*$. $e^* \step v^*$, so $\tau^{*\prime} = \tau^* = \tau_{elem}^*$.

    \item \Rulename{ADD} \footnote{For these three rules, terms of the form $v_n$ are better expressed as numeric literals.}
        Then $e^* = v_1~ v_2~ +$ and $e^* \step v_3$, where $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
        By \Rulename{T-5}, $\etyped{v_1~ v_2~ +}{\tau}$, and by the appropriate choice of \Rulename{T-1}, \Rulename{T-2}, \Rulename{T-3}, or \Rulename{T-4}, $\etyped{v_3}{\tau}$., so $\tau^* = \tau^{*\prime} = \tau$.

    \item \Rulename{MUL}
        Omitted for brevity on the grounds of being mechanically identical to the above case.

    \item \Rulename{EQUIV}
        Then $e^* = v_1~ v_2~ ==$ where $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
        By \Rulename{T-8}, $\tau^* = \kwInt$.
        $e^* \step \kwnumlit{n}{\kwInt}$, and by \Rulename{T-1}, $\etyped{\kwnumlit{n}{\kwInt}}{\kwInt}$, so $\tau^{*\prime} = \tau^* = \kwInt$.

    \item \Rulename{TERN-1}, \Rulename{TERN-2}
        Then $e^* = \kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2}$ and $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau$.
        By \Rulename{T-7}, $\etyped{e^*}{\tau}$, so $\tau^* = \tau$.
        Either $e^* \step v_1$ or $e_* \step v_2$, so in either case, $\tau^{*\prime} = \tau$.
\end{itemize}

\end{proof}

% \subsubsection{Lemma - Convergence}
% Part 1: $\tau^* \tstep \tau^{*\prime}$ is no longer possible when $\tau_i \prim ~ \forall~ \tau_i \in \tau^*$.
% \begin{proof}
% Proof by exhaustion with all rules involving $\prim$ types not stepping. To values
% \end{proof}
% Part 2: $\tau^* \tstep \tau^{*\prime}$ will eventually step to a $\tau^{*\prime}$ such that $\tau_i \prim ~ \forall~ \tau_i \in \tau^{*\prime}$.
% \begin{proof}
% This is provable by the fact that the language does not feature recursive types and the only form of recursion in the language is via the fix operator and $\kwType{}$ operator does no have to follow recursive paths in order to get it's type (Type Rules \Rulename{Macro}, \Rulename{Fix}). The other rules are trivial (ie - step to a $\prim$) and/or irrelevant to the assertion
% \end{proof}
